Case 2 Lab
========================================================
### Executive Summary  
This tutorial illustrates the implentations of clustering and association rule mining in (R), which will help you finish Case Study 2.

****
<a id="content"></a> Content 
-------------------
### [Data] (#data)   
### [RODBC](#rodbc)   
### [Clustering](#cluster)   
### [Association Rules](#arules)  
****

### <a id="data"></a>Cincinnati Zoo Data 
This lab uses the Cincinnati Zoo Data, which is from former MSQA student Ben Cofie's Master's thesis. A copy of his thesis can be found on BB.

### <a id="rodbc"></a>RODBC package
The [RODBC](http://cran.r-project.org/web/packages/RODBC/RODBC.pdf) package provides access to databases (including Microsoft Access and Microsoft SQL Server) through an ODBC interface.
The primary functions are given below. 

Function |Description
------------|---------------
odbcConnect(dsn, uid="", pwd="") | Open a connection to an ODBC database
sqlFetch(channel, sqtable)	 | Read a table from an ODBC database into a data frame
sqlQuery(channel, query)	| Submit a query to an ODBC database and return the results
sqlSave(channel, mydf, tablename = sqtable, append = FALSE) | 	Write or update (append=True) a data frame to a table in the ODBC database
sqlDrop(channel, sqtable)	 | Remove a table from the ODBC database
close(channel)	| Close the connection

****
### <a id="cluster"></a>Clustering
Data I/O and Some Preprocessing
```{r,eval=F}
#Import raw data from excel
library(RODBC)
channel <- odbcConnectExcel("D:/qry_Food_by_Month.xls")
ZooFood <- sqlFetch(channel, "qry_Food_by_MonthEdited")
odbcClose(channel)
```
**Note:**  *odbcConnectExcel* function is only usable with 32-bit Windows. To change setting in Rstudio, go to Tools -> Options -> General R Versions: select 32 bit R version. You may need to restart Rstudio to make the change take effect.


```{r}
load("I:/BANA7047/Case 2/case2 RWSpace.RData")
```

```{r}
#Delete missing rows with missing data
ZooFood<-na.omit(ZooFood)
#Exclude column 1 which contains text from data matrix
ZooFood.use<- ZooFood [,-1]
#re-name the rows
dimnames(ZooFood.use)[[1]] <- ZooFood[,1]
#Scale data using the column means.
ZooFood.use<-scale(ZooFood.use) #Scale Data to have zero mean and unit variance for each column
```


Determine number of clusters. Here a simple within group sum of squares method is used. For more sophiscated methods, see for example [blog](http://blog.echen.me/2011/03/19/counting-clusters/), or [course notes](http://www.bx.psu.edu/old/courses/bx-fall04/How_Many_Clusters.pdf).
```{r}
# Determine number of clusters
wss <- (nrow(ZooFood.use)-1)*sum(apply(ZooFood.use,2,var))
for (i in 2:12) wss[i] <- sum(kmeans(ZooFood.use,
                                     centers=i)$withinss)
plot(1:12, wss, type="b", xlab="Number of Clusters",ylab="Within groups sum of squares")
```

Use k-means method for clustering and plot results.
```{r}
# K-Means Cluster Analysis
fit <- kmeans(ZooFood.use, 5) #5 cluster solution
#Display number of clusters in each cluster
table(fit$cluster)
#Plot cluster in kmeans
install.packages("fpc")
library(fpc)
plotcluster(ZooFood.use, fit$cluster)
#See exactly which item are in 1st group
ZooFood$NickName[fit$cluster==1]
# get cluster means for scaled data
aggregate(ZooFood.use,by=list(fit$cluster),FUN=mean)
#or alternatively, use the output of kmeans
fit$centers
#get cluster means for raw data
Ageish<-aggregate(ZooFood[,-c(1,1)],by=list(fit$cluster),FUN=mean)
Ageish
```

```{r}
#Wards Method or Hierarchical clustering
#Calculate the distance matrix
ZooFood.dist=dist(ZooFood.use)
#Obtain clusters using the Wards method
ZooFood.hclust=hclust(ZooFood.dist, method="ward")
plot(ZooFood.hclust)
#Cut dendrogram at the 3 clusters level and obtain cluster membership
groupZooFood.3 = cutree(ZooFood.hclust,k=3)
table(groupZooFood.3)
#See exactly which item are in third group
ZooFood$NickName[groupZooFood.3==3]
#get cluster means for raw data
aggregate(ZooFood[,-c(1,1)],by=list(groupZooFood.3),FUN=mean)
# Centroid Plot against 1st 2 discriminant functions
#Load the fpc library needed for plotcluster function
library(fpc)
#plotcluster(ZooFood, fit$cluster)
plotcluster(ZooFood.use, groupZooFood.3)
```

[go to top](#content)
****
### <a id="arules"></a>Association Rules
Association Rules is a popular and well researched method for discovering
interesting relations between variables in large databases. [*arules*](http://cran.r-project.org/web/packages/arules/arules.pdf) package in R provides a basic infrastructure for creating and manipulating input data sets and for analyzing the resulting itemsets and rules.

Data I/O and summary of transaction data
```{r,eval=FALSE}
#Import raw data from excel
library(RODBC)
channel <- odbcConnectExcel("I:/BANA7047/Case 2/food_4_association.xls")
TransFood <- sqlFetch(channel, "Sheet1")
odbcClose(channel)
```

```{r}
load("I:/BANA7047/Case 2/case2 RWSpace.RData")

TransFood.use<- TransFood[,-c(1,1)]

#Load the arules library
library(arules)

#Transform raw data into matrix form
TransFood.use<-as(as.matrix(TransFood.use),"transactions")
#TransFood.use <- as.matrix(TransFood.use)

#run summary report
summary(TransFood.use)
```
*as()* function coerce the dataset into transaction data type for association rules mining. *summary()* displays the most frequent items in the data set, information about the transaction
length distribution and that the data set contains some extended transaction information.
We see that the data set contains transaction IDs. This additional information can be used for analyzing the data
set. 

To find the very long transactions we can use the *size()* and select very long transactions (containing more than 10 items).
```{r}
#
transactionInfo(TransFood.use[size(TransFood.use) > 10])
```

To see which items are important in the data set we can use the *itemFrequencyPlot()*. To reduce the number of items, we only plot the item frequency for items with a support greater than 10%. The label size is reduced with the parameter *cex.names*.
```{r}
#
itemFrequencyPlot(TransFood.use, support = 0.1, cex.names=0.8)

```


Use *apriori()* algorithm to find all rules (the default association type for
*apriori()*) with a minimum support of 1% and a confidence of 0.6.
```{r}
# Run the apriori algorithm
basket_rules <- apriori(TransFood.use,parameter = list(sup = 0.003, conf = 0.7,target="rules"))

summary(basket_rules)
```
Recall from class:
* **Support**: The support of an itemset X or $latex supp(X)$ is defined as the proportion of
transactions in the data set which contain the itemset. In the zoo data, the support for the rules is relatively low, with a maximum support of no more than 3%. 
* **Confidence**: The confidence of a rule is defined as $latex conf( X\Rightarrow Y) = supp( X \cup Y )/supp(X)$. For example, the
rule {milk, bread} $latex \Rightarrow$ {butter} has a confidence of 0.5,
which means that for 50% of the transactions containing milk and bread the rule is correct.
Confidence can be interpreted as an estimate of the conditional probability P(Y |X), the probability of
finding the RHS of the rule in transactions under the condition that these transactions also
contain the LHS. Association rules are required to satisfy both a minimum support and a minimum confidence
constraint at the same time.

* **Lift**: Life is a popular measure of to filter or rank found rules. The lift of a rule is defined as $latex lift(X \Rightarrow Y ) = supp(X \cup Y )/(supp(X)supp(Y ))$. Lift can be interpreted as the deviation of the support of the whole rule from the support expected
under independence given the supports of the LHS and the RHS. Greater lift values indicate
stronger associations.
```{r}
# Check the generated rules using inspect
inspect(basket_rules)
```

As typical for association rule mining, the number of rules found is huge. To analyze these
rules, for example, subset() can be used to produce separate subsets of rules. Now find the subset of rules that has 3 or more length (lhs+rhs).
```{r}
#Basket rules of size greater than 2
inspect(subset(basket_rules, size(basket_rules)>2))

#Basket rules of size greater than 3
inspect(subset(basket_rules, size(basket_rules)>3))

```
There are no rules that have more than 3 items in total (lhs+rhs).

Now find the subset rules that has "Ice Cream ConeFood" in the right hand side. Here we require *lift* measure exceeds 1.2.
```{r}
icecream.rhs <- subset(basket_rules, subset = rhs %in% "Ice Cream ConeFood" & lift>1.2)
```
Now inspect the subset rules
```{r}
inspect(icecream.rhs)
```

Now find the subset rules that has "Bottled WaterFood" in the left hand side.
```{r}
water.lhs <- subset(basket_rules, subset = lhs %in% "Bottled WaterFood" & lift>1.2)
```
Now inspect the subset rules
```{r}
inspect(water.lhs)
```

[go to top](#content)